\vspace{-2.0em} Move Schedules: \\ Fast persistence computations in sparse dynamic settings\thanks{This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}\vspace{-0.5em}

### 2021-08-05 move_schedules.tex
Beginning the bridge to describe the detection example better

There are several approaches one can use to compute $\mathcal{T}$, the simplest being to run Algorithm~\ref{alg:reduce} independently on the 1-D filtration induced by the duals of some set of points (e.g. the barycenters) lying in the interior of the 2-cells of $\mathcal{A}(M)$.  
The approach taken by~\cite{lesnick2015interactive} is to use the $R = DV$ decomposition computed at some adjacent 2-cell $e \in \mathcal{A}(M)$ to speed up the computation of an adjacent cell $e' \in \mathcal{A}(M)$. More explicitly, define the \emph{dual graph} of $\mathcal{A}(M)$ to be the undirected graph $G$ which has a vertex for every 2-cell $e \in \mathcal{A}(M)$ and an edge for each adjacent pair of cells $e, e' \in \mathcal{A}(M)$.
Each vertex in $G$ is associated with a barcode template $\mathcal{T}^e$, and the computation of $\mathcal{T}$ now reduces to computing a path $\Gamma$ on $G$ which visits each vertex at least once. To minimize the computation time, assume the $n$ edges of $G$ are endowed with non-negative weights $W = w_1, w_2, \dots, w_n$ whose values $w_i \in \mathbb{R}_{+}$ represent some notion of distance. The optimal path $\Gamma^\ast$ is the one of minimal length with respect to $W$ which visits every vertex of $G$ at least once. There is a known $\frac{3}{2}$-approximation that can be computed efficiently which reduces the problem to the traveling salesman problem on a metric graph~\cite{christofides1976worst}. In the next paragraph, we include a small example to illustrate many of the concepts and definitions discussed so far below. 
\\
\\
\textbf{Noisy circle example}
Consider a small set of points distributed noisily around $S^1$ that contains a few strong outliers, shown on the left side of Figure~\ref{fig:bifiltration_ex2}. Filtering this data set with respect to the Rips parameter and the complement of a kernel density estimate yields a bifiltration whose dimension function and bigraded Betti numbers are shown in the middle figure. The gray areas indicate the region where the dimension function is constant. On the right side, the space dual to the space in the middle is shown. The black lines represent the lines of $\mathcal{A}(M)$, the blue dashed-lines the edges of the dual graph $\mathcal{G}$, and the barycenters of the 2-cells in $\mathcal{A}(M)$ depicting where the barcodes templates $\mathcal{T}^e$ are computed are drawn as orange points. 
\begin{figure}[t]
	\includegraphics[width=0.98\textwidth]{bifiltration_ex2}
	\caption{Bipersistence example on an $8 \times 8$ coarsened grid. On left, the input data, colored by density. In the middle, the bigraded Betti numbers $\beta_0(M)$ and $\beta_1(M)$ (green and red, respectively), the Hilbert function (gray), and a line $L$ emphasizing the persistence of features with high density. On the right, the line arrangement $\mathcal{A}(M)$ lying in the dual space $D_p(\alpha)$ derived from the $\beta(M)$. }
	\label{fig:bifiltration_ex2}
\end{figure}
The shortest path $\Gamma^\ast$ spanning the vertices of $G$ is shown by the solid rainbow colored lines, with the traversal beginning in the top-left cell (at the red edge) and ending at the bottom-right cell (the indigo edge). Traversing $\Gamma^\ast$ can be done via a simple depth-first search, though note that a $R=DV$ decomposition needs to be stored at each degree-3 or higher vertex in $\Gamma^\ast$ to avoid redundant computations as the traversal recurses. 
\\

@@

 \begin{figure}[t]
	\includegraphics[width=0.98\textwidth]{natural_images}
	\caption{Bipersistence example of natural images data set on an $12 \times 16$ coarsened grid. On the left, a projection the full data set is shown, along with the 15 landmark patches. In the middle, the bigraded Betti numbers, the Hilbert function, and a chosen line $L$ are shown for the coarsened grid. As before, the $0/1/2$ dimension bigraded Betti numbers are shown in green/red/yellow, respectively. The blue regions highlights where $\mathrm{dim}(M) = 5$. On the right, the persistence diagram of $L$, with larger points indicating a multiplicity of 2. Observe five persistent features are revealed in both the middle and right plots, matching $\beta_1$ of the three-circle model. }
	\label{fig:patch_data_dgm}
\end{figure}

A common hypothesis is that high dimensional data tend to lie in the vicinity of an embedded, low dimensional manifold or topological space. An exemplary demonstration of this is given in~\cite{lee2003nonlinear}, who explored the space of $3 \times 3$ high-contrast patches extracted from Hans van Hateren's still image collection~\cite{hateren_schaaf_1998}, which consists of approximately 4,000 monochrome images\footnote{See \url{http://bethgelab.org/datasets/vanhateren/} for details on the image collection.} depicting various areas outside around Groningen (Holland). The study of how empirical data extracted from natural images is distributed with respect to predicted spaces and manifolds is motivated by applications in sparse coding and efficient image representations. In particular, Lee et al.~\cite{lee2003nonlinear} was interested in analyzing the empirical probability distribution of high-contrast $3 \times 3$ patches of natural images, in pixel-space, towards understanding whether there were any clear qualitative differences the distributions of patches extracted from from images of different modalities, e.g. optical versus range images. 
Experiments conducted by Carllsson et al.~\cite{carlsson2008local} using persistent homology confirmed that the distribution of high-contrast $3 \times 3$ patches is indeed well-approximated by a Klein bottle---around 60\% of the high-contrast patches from the still image data set are concentrated within a small neighborhood around a model of a Klein bottle which accounts for $\approx$ 21\% of the volume $S^7$, compared to the $\approx$ 84\% of volume of $S^7$ accounted for by the entire data set. Along a similar vein but from a sparse coding perspective, Perea et al.~\cite{perea2014klein} introduced dictionary learning framework for estimating and representing the distribution of patches from texture images.

If one was not aware of the analysis done by~\cite{lee2003nonlinear, hateren_schaaf_1998, carlsson2008local, perea2014klein}, there is no specific reason apriori to assume the data are well-approximated by the Klein bottle. Consider the (coarsened) fibered barcode shown in Figure~\ref{fig:patch_data_dgm} which was computed from a standard Rips/codensity bifiltration on a constructed on a representative sample of the image data from~\cite{hateren_schaaf_1998} (how we obtained this sample is described further down). Upon inspection of the bigraded Betti number and the dimension function, one finds that a large area of dimension function is constant (highlighted as the blue portion in the middle of Figure~\ref{fig:patch_data_dgm}), wherein $\mathrm{dim}(M) = 5$. This indicates

>>>>

There are several approaches one can use to compute $\mathcal{T}$, the simplest being to run Algorithm~\ref{alg:reduce} independently on the 1-D filtration induced by the duals of some set of points (e.g. the barycenters) lying in the interior of the 2-cells of $\mathcal{A}(M)$.  
The approach taken by~\cite{lesnick2015interactive} is to use the $R = DV$ decomposition computed at some adjacent 2-cell $e \in \mathcal{A}(M)$ to speed up the computation of an adjacent cell $e' \in \mathcal{A}(M)$. More explicitly, define the \emph{dual graph} of $\mathcal{A}(M)$ to be the undirected graph $G$ which has a vertex for every 2-cell $e \in \mathcal{A}(M)$ and an edge for each adjacent pair of cells $e, e' \in \mathcal{A}(M)$.
Each vertex in $G$ is associated with a barcode template $\mathcal{T}^e$, and the computation of $\mathcal{T}$ now reduces to computing a path $\Gamma$ on $G$ which visits each vertex at least once. To minimize the computation time, assume the $n$ edges of $G$ are endowed with non-negative weights $W = w_1, w_2, \dots, w_n$ whose values $w_i \in \mathbb{R}_{+}$ represent some notion of distance. The optimal path $\Gamma^\ast$ is the one of minimal length with respect to $W$ which visits every vertex of $G$ at least once. There is a known $\frac{3}{2}$-approximation that can be computed efficiently which reduces the problem to the traveling salesman problem on a metric graph~\cite{christofides1976worst}. In the next paragraph, we include a small example to illustrate many of the concepts and definitions discussed so far below. 
\\
\\
\textbf{Noisy circle example}
Consider a small set of points distributed noisily around $S^1$ that contains a few strong outliers, shown on the left side of Figure~\ref{fig:bifiltration_ex2}. Filtering this data set with respect to the Rips parameter and the complement of a kernel density estimate yields a bifiltration whose dimension function and bigraded Betti numbers are shown in the middle figure. The gray areas indicate the region where the dimension function is constant. In particular, the light gray area of the middle plot shows where $\mathrm{dim}_1(M) = 1$, indicating where in the multifiltration a persistent 1-cycle was detected.
On the right side, the space dual to the space in the middle is shown. The black lines represent the lines of $\mathcal{A}(M)$, the blue dashed-lines the edges of the dual graph $\mathcal{G}$, and the barycenters of the 2-cells in $\mathcal{A}(M)$ depicting where the barcodes templates $\mathcal{T}^e$ are computed are drawn as orange points. 
\begin{figure}[t]
	\includegraphics[width=0.98\textwidth]{bifiltration_ex2}
	\caption{Bipersistence example on an $8 \times 8$ coarsened grid. On left, the input data, colored by density. In the middle, the bigraded Betti numbers $\beta_0(M)$ and $\beta_1(M)$ (green and red, respectively), the Hilbert function (gray), and a line $L$ emphasizing the persistence of features with high density. On the right, the line arrangement $\mathcal{A}(M)$ lying in the dual space $D_p(\alpha)$ derived from the $\beta(M)$. }
	\label{fig:bifiltration_ex2}
\end{figure}
The shortest path $\Gamma^\ast$ spanning the vertices of $G$ is shown by the solid rainbow colored lines, with the traversal beginning in the top-left cell (at the red edge) and ending at the bottom-right cell (the indigo edge). Traversing $\Gamma^\ast$ can be done via a simple depth-first search, though note that a $R=DV$ decomposition needs to be stored at each degree-3 or higher vertex in $\Gamma^\ast$ to avoid redundant computations as the traversal recurses. 
\\

@@

 \begin{figure}[t]
	\includegraphics[width=0.98\textwidth]{natural_images}
	\caption{Bipersistence example of natural images data set on an $12 \times 16$ coarsened grid. On the left, a projection the full data set is shown, along with the 15 landmark patches. In the middle, the bigraded Betti numbers, the Hilbert function, and a chosen line $L$ are shown for the coarsened grid. As before, the $0/1/2$ dimension bigraded Betti numbers are shown in green/red/yellow, respectively. The blue regions highlights where $\mathrm{dim}(M) = 5$. On the right, the persistence diagram of $M$ restricted to $L$, with larger points indicating a multiplicity of 2. Observe five persistent features are revealed in both the middle and right plots, matching $\beta_1$ of the three-circle model. }
	\label{fig:patch_data_dgm}
\end{figure}

A common hypothesis is that high dimensional data tend to lie in the vicinity of an embedded, low dimensional manifold or topological space. An exemplary demonstration of this is given in~\cite{lee2003nonlinear}, who explored the space of $3 \times 3$ high-contrast patches extracted from Hans van Hateren's still image collection~\cite{hateren_schaaf_1998}, which consists of approximately 4,000 monochrome images\footnote{See \url{http://bethgelab.org/datasets/vanhateren/} for details on the image collection.} depicting various areas outside around Groningen (Holland). Motivated by applications to sparse coding applications, Lee et al.~\cite{lee2003nonlinear} was interested in how these natural image patches were distributed, in pixel-space, with respect to predicted spaces and manifolds. For example, one of the motivating questions of their research was whether there were any clear qualitative differences the distributions of patches extracted from from images of different modalities, e.g. optical versus range images. A major result from their work is that the majority of high-contrast image patches are concentrated around a 2-dimensional submanifold embedded in $S^7$.
Supplemental experiments conducted by Carllsson et al.~\cite{carlsson2008local} using persistent homology confirmed that the distribution of high-contrast $3 \times 3$ patches is indeed well-approximated by a Klein bottle---around 60\% of the high-contrast patches from the still image data set are concentrated within a small neighborhood around a model of a Klein bottle which accounts for $\approx$ 21\% of the volume $S^7$, compared to the $\approx$ 84\% of volume of $S^7$ accounted for by the entire data set. Along a similar vein, in the context of sparse coding, Perea et al.~\cite{perea2014klein} introduced a dictionary learning framework for estimating and representing the distribution of patches from texture images.

If one was not aware of the analysis done by~\cite{lee2003nonlinear, hateren_schaaf_1998, carlsson2008local, perea2014klein}, it is not immediately clear a priori that the Klein bottle model is indeed an excellent candidate for capturing the non-linearity of the data. Nonetheless, as we have mentioned, applications of persistent homology detecting the presence of a topological space persistent homology is designed to reveal. Consider the (coarsened) fibered barcode shown in Figure~\ref{fig:patch_data_dgm} which was computed from a standard Rips/codensity bifiltration on a constructed on a representative sample of the image data from~\cite{hateren_schaaf_1998} (how we obtained this sample is described further down). Upon inspection of the bigraded Betti number and the dimension function, one finds that a large area of dimension function is constant (highlighted as the blue portion in the middle of Figure~\ref{fig:patch_data_dgm}), wherein $\mathrm{dim}(M) = 5$. This indicates
